Concurrent Programming with Harmony

### 2020-07-30 book/paper.tex
Improved monitor chapter

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}%
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
the baton to some process $q$ requiring the baton (assuming there even
is such a process).  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point, any process that
was notified will have a chance to obtain the baton.

@@

On the left is a Hoare monitor.  When a process leaves the monitor
(let's call that process $p$), it must be in the bathroom.
It can choose to either let in a process
in one of the bedrooms or a process in the hall, assuming there are any.
To make the choice, $p$ checks for each bedroom if its condition holds
and if there are processes in the bedroom.
If there are such bedrooms, $p$ selects one and one of its processes into
the bathroom.  Let's call that process $q$.  Importantly, when $q$ enters
the bathroom, the condition that $q$ was waiting for still holds.

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When process $p$ leaves the bathroom,
if also checks each of the bedrooms to see if there is a process $q$ that
can run.  But instead of letting $q$ go straight into the bathroom, $q$
goes into the hall, joining any other processes that are also waiting to
enter the bathroom.  Finally, when $p$ has left the bathroom, one of the
processes in the hall is let into the bathroom.  But that may not be $q$.
Assuming every process eventually leaves the bathroom and the system is
fair, eventually $q$ will enter the bathroom.  However, it is no longer
guaranteed that its condition holds.  Therefore, the first thing $q$ must
do is check, again, to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.

Mesa monitors provide another important option:
when process $p$ leaves the bathroom, it can choose to let any number of
processes in the bedrooms into the hall.  It can do so by calling
\texttt{notify} repeatedly, each time letting one process in one of the
bedrooms into the hall.  Using \texttt{notify} on an empty
bedroom is considered a no-op in Mesa.  (Hoare monitors do not allow
using \texttt{signal} on an empty bathroom.) Mesa also provides an
operation called \texttt{notifyAll}
\index{notifyAll}%
(aka \texttt{broadcast)})
\index{broadcast}%
that, when applied to
a bedroom, lets all processes in that bedroom into the hall.
This is not possible with Hoare monitors: when needing to wake up a set
of processes, only one can be signaled at a time.  Each of those processes
then has to signal the next until all processes have been signaled.

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular, adopted by all major
programming languages.
That said, few programming languages provide full support for monitors.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.  Java also supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is through libraries.

@@

\begin{figure}
\begin{code}
\harmonysource{RWcv}
\end{code}
\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock implementation using Mesa-style condition variables.}
\label{fig:RWcv}
\end{figure}

@@

>>>>

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a process can wait on.
Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}%
When a process $p$ invokes \texttt{notify}, it does not immediately pass 
the baton to some process $q$ requiring the baton (assuming there even
is such a process).  Instead, $p$ holds onto
the baton until it leaves the monitor.  At that point, any process that
was notified will have a chance to obtain the baton, but they compete
with other processes also trying to obtain the baton.

@@

On the left is a Hoare monitor.
When some process $p$ is in the bathroom, it can invoke \textbf{signal}
on some non-empty bedroom associated with a condition that holds.
This select a process $q$ in that bedroom.
As a result of invoking \textbf{signal}, $p$ goes back into the hall
and $q$ enters the bathroom.
Importantly, when $q$ enters the bathroom, the condition that $q$ was
waiting for is guaranteed to hold.

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When process $p$ invokes \textbf{notify} on a bedroom, it
selects a process $q$ in that bedroom, if any.
Process $q$ goes into the hall,
joining any other processes that are also waiting to
enter the bathroom.
If there are no processes in that bedroom, \textbf{notify} is a no-op.
Importantly, in either case $p$ remains in the bathroom.
Finally, when $p$ leaves the bathroom, one of the
processes in the hall can enter the bathroom.  But that may not be $q$.
Assuming every process eventually leaves the bathroom and the system is
fair, eventually $q$ will enter the bathroom.  However, it is no longer
guaranteed that the condition it was waiting for still holds.
Therefore, the first thing $q$ must
do is check, again, to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.

Besides notifying processes, Mesa monitors provide another important option:
it can choose to let any number of processes in the bedrooms into the hall.
It can do so by calling
\texttt{notify} repeatedly, each time letting one process in the specified
bedroom into the hall.  Mesa also provides an
operation called \texttt{notifyAll}
\index{notifyAll}%
(aka \texttt{broadcast)})
\index{broadcast}%
that, when applied to
a bedroom, lets all processes in that bedroom into the hall.
This is not possible with Hoare monitors: when needing to wake up a set
of processes, only one can be signaled at a time.  Each of those processes
then has to signal the next in a baton-passing style
until all processes have been signaled.

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular, adopted by all major
programming languages.
That said, few programming languages provide full support for monitors.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.
In addition, Java supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is through libraries.

@@

\begin{figure}
\begin{code}
\harmonysource{RWcv}
\end{code}
\caption{\harmonylink{code/RWcv.hny} Reader/Writer Lock using Mesa-style condition variables.}
\label{fig:RWcv}
\end{figure}

@@

While Hoare moniters embody the split binary semaphore approach,
Mesa monitors are much closer to the busy-waiting approach.
When comparing the busy-waiting version of reader/writer locks in \autoref{fig:rwbusy}
and the Mesa version in \autoref{fig:RWcv}, you will see important similarities.
Busy-waiting is ``fixed'' by placing a \textbf{wait} statement in the waiting loop.
The Mesa code then has to be careful to invoke \textbf{notify} or \textbf{notifyAll}
in the right places.
